---
title: "Lab 3: MAUP in Practice — Tessellations and Point-in-Polygon"
subtitle: "National Dam Inventory (NID): From Counts to Local Clustering"
author: "Maisie Moline"
format:
  html:
    code-fold: true
    toc: true
    include-before-body: ../slides/header.html
    include-after-body: ../slides/footer-annotations.html
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| include: false
knitr::opts_chunk$set(
  fig.width  = 8,
  message    = FALSE,
  warning    = FALSE,
  comment    = "",
  cache      = FALSE,
  fig.retina = 3
)
```

# Background

In this lab we will explore the impacts of tessellated surfaces and the
modifiable areal unit problem (MAUP) using the National Dam Inventory
maintained by the United States Army Corps of Engineers. The work
requires writing reusable functions and careful consideration of feature
aggregation/simplification, spatial joins, and data visualization. The
end goal is to visualize the distribution of dams and their purposes
across the country.

**DISCLAIMER**: This lab can be computationally intensive. For some
steps you may be processing hundreds of millions of relations; run code
chunk-by-chunk during development rather than knitting the whole
document repeatedly. Your final knit may take a few minutes. Be proud —
the report will summarize analyses over very large geometric relations.

------------------------------------------------------------------------

This lab covers 4 main skills:

1.  **Tessellating surfaces** to discretize space
2.  **Geometry simplification** to expedite expensive intersections
3.  **Writing functions** to automate repetitive reporting and mapping
    tasks
4.  **Point-in-polygon counts** to aggregate point data

## Graduate-Level Expectations

In addition to producing correct code and outputs, your submission
should:

-   **Justify methodological choices**: Why did you choose a particular
    tessellation `keep` percentage for simplification? How did you
    balance computational efficiency with map detail?
-   **Document performance impacts**: Measure and report computation
    time and geometric accuracy changes resulting from simplification.
    How do results differ?
-   **Apply spatial reasoning**: Use spatial predicates (e.g.,
    `st_touches`, `st_contains`) to identify and describe geometric
    relationships in the dam dataset that raw counts cannot reveal.
-   **Compare across methods**: When multiple tessellation approaches
    yield different clustering patterns, explain which pattern better
    supports real-world dam management decisions and why.
-   **Connect to lecture**: Explicitly cite the MAUP problem,
    simplification algorithms (Douglas-Peucker vs. Visvalingam), and
    spatial predicate definitions from week 3 lectures in your analysis.

### Libraries

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false

# Data Science
library(tidyverse) # data wrangling
library(sf)        # Working with vector data
library(rmapshaper)# Simplify geometries
library(units)     # manage your units
library(dplyr)
library(sf)

# Data
# remotes::install_github("mikejohnson51/AOI") # get US counties
library(AOI) # county boundaries

# Visualization
library(gghighlight) # ggplot conditional highlighting
library(knitr) # table generation
library(kableExtra) # making tables pretty
```

------------------------------------------------------------------------

# Question 1:

Here we will prepare five tessellated surfaces from CONUS and write a
function to plot them in a descriptive way.

### Step 1.1

First, we need a spatial file of CONUS counties. For future area
calculations we want these in an equal area projection (`EPSG:5070`).

To achieve this:

-   get an `sf` object of US counties
    (`AOI::aoi_get(state = "conus", county = "all")`)

-   transform the data to `EPSG:5070`

```{r}

county_conus<-AOI::aoi_get(state = "conus", county = "all")

county_conus<-county_conus %>% st_transform(5070) %>% st_make_valid()
```

### Step 1.2

For triangle based tessellations we need point locations to serve as our
"anchors".

To achieve this:

-   generate county centroids using `st_centroid`

-   Since, we can only tessellate over a feature we need to *combine* or
    *union* the resulting 3,108 `POINT` features into a single
    `MULTIPOINT` feature

-   Since these are point objects, the difference between union/combine
    is mute

```{r}

centroids<- county_conus %>% st_centroid() %>% st_combine()
```

### Step 1.3

Tessellations/Coverage's describe the **extent** of a region with
geometric shapes, called **tiles**, with *no* overlaps or gaps.

**Tiles** can range in *size*, *shape*, *area* and have different
methods for being created.

Some methods generate triangular tiles across a set of defined points
(e.g. `Voronoi` and Delaunay triangulation)

Others generate equal area tiles over a known extent (`st_make_grid`)

For this lab, we will create surfaces of CONUS using using 4 methods, 2
based on an extent and 2 based on point anchors:

**Tessellations** :

-   `st_voronoi`: creates a Voronoi tessellation

-   `st_triangulate`: triangulates a set of points (not constrained)

**Coverages**:

-   `st_make_grid`: creates a square grid covering the geometry of an sf
    or sfc object

-   `st_make_grid(square = FALSE)`: creates a hexagonal grid covering
    the geometry of an sf or sfc object

-   The side of coverage tiles can be defined by a cell resolution or a
    specified number of cell in the X and Y direction

------------------------------------------------------------------------

For this step:

-   Make a voroni tessellation over your county centroids (`MULTIPOINT`)
-   Make a triangulated tessellation over your county centroids
    (`MULTIPOINT`)
-   Make a gridded coverage with n = 70, over your counties object
-   Make a hexagonal coverage with n = 70, over your counties object

In addition to creating these 4 coverage's we need to add an ID to each
*tile*.

To do this:

-   add a new column to each tessellation that spans from `1:n()`.

-   Remember that **ALL** tessellation methods return an `sfc`
    `GEOMETRYCOLLECTION`, and to add attribute information - like our
    ID - you will have to coerce the `sfc` list into an `sf` object
    (`st_sf` or `st_as_sf`)

Last, we want to ensure that our surfaces are topologically
valid/simple.

-   To ensure this, we can pass our surfaces through `st_cast`.

-   Remember that casting an object explicitly (e.g.
    `st_cast(x, "POINT")`) changes a geometry

-   If no output type is specified (e.g. `st_cast(x)`) then the cast
    attempts to simplify the geometry.

-   If you don't do this you might get unexpected "TopologyException"
    errors.

```{r}

#Step: voroni tessellation over county cent mp

voronoi_county<-st_voronoi(centroids) %>% st_cast() %>% st_as_sf() %>% mutate(id=1:n())


#triangulated tessellation over your county centroids 

tri_county<-st_triangulate(centroids) %>% st_cast() %>% st_as_sf() %>% mutate(id=1:n())

# gridded coverage with n = 70, over your counties object

county_grid<-st_make_grid(county_conus,n=70) %>%st_cast() %>%  st_as_sf() %>% mutate(id=1:n())

#hex coverage
county_hex<-st_make_grid(county_conus,n=70,square = F) %>% st_cast() %>% st_as_sf() %>% mutate(id=1:n())

county_id<-
st_sf(county_conus) %>% mutate(id = 1:n())
```

### Step 1.4

If you plot the above tessellations you'll see that the **Voronoi**,
**Delaunay triangulation**, **hexagon**, and **square grids** all
produce regions far beyond the boundaries of CONUS.

We need to cut these boundaries to CONUS borders.

To do this, we need two different approaches: - **`st_intersection()`**
for Voronoi and Delaunay (these tessellations create geometries that
extend infinitely/far beyond the study area; intersection clips them at
the exact boundary) - **`st_filter()`** for hexagon and square grids
(these regular grids are already designed to fit; filtering keeps only
the tiles that intersect CONUS)

Either way, we'll need a geometry of CONUS to serve as our clipping
feature. We can get this by unioning our existing county boundaries.

```{r, eval = TRUE}
conus_border<-st_union(county_conus)


```

### Step 1.5

With a single feature boundary, we must carefully consider the
complexity of the geometry. Remember, the more points our geometry
contains, the more computations needed for spatial predicates our
differencing. For a task like ours, we do not need a finely resolved
coastal boarder.

To achieve this:

-   Simplify your unioned border using the Visvalingam algorithm
    provided by `rmapshaper::ms_simplify`.

-   Choose what percentage of vertices to retain using the `keep`
    argument and work to find the highest number that provides a shape
    *you* are comfortable with for the analysis:

```{r, eval = TRUE}
conus_simp<-rmapshaper::ms_simplify(conus_border,keep = 0.1)

st_is_valid(conus_simp)

# Check: Does it still contain all of CONUS?
st_bbox(conus_simp)  # Should cover roughly [-125, 24] to [-66, 49]

# Check: Visual comparison
plot(conus_border, main = "Original")
plot(conus_simp, main = "Simplified")

```

::: callout-important
## Validation Checkpoint (Optional)

After simplification, verify the boundary is still valid and covers all
dams:

```{r}
#| eval: false
# Check: Is the simplified boundary valid?
st_is_valid(conus_simp)

# Check: Does it still contain all of CONUS?
st_bbox(conus_simp)  # Should cover roughly [-125, 24] to [-66, 49]

# Check: Visual comparison
plot(conus_border, main = "Original")
plot(conus_simp, main = "Simplified")
```
:::

::: callout-note
## Simplification: Boundary Only, Not Tessellation

**Important distinction**: This step simplifies the **CONUS boundary
container** we use for clipping the tessellated surfaces. Simplification
affects *only* the geometry of the clipping operation — it speeds up the
intersection computation by reducing boundary vertices. Your
tessellation tiles themselves remain **high-resolution** and unchanged.
You are not simplifying the tessellation itself, only the container we
clip with.
:::

-   Once you are happy with your simplification, use the `mapview::npts`
    function to report the number of points in your original object, and
    the number of points in your simplified object.

-   How many points were you able to remove? What are the consequences
    of doing this computationally?

    > I was able to get the points from 11292 to 1145, a removal of
    > 10,147 points. A positive consequence is that computations will
    > run quicker now. A negative consequence of this is that topology
    > is lost which can cause overlapping or disconnected geometries.
    > This can impact detail seen on coastlines and near borders.

```{r}
library(mapview)
npts(conus_simp)
npts(conus_border)
```

-   Finally, crop **all four grids** to the CONUS boundary—but using
    different methods:
    -   **Voronoi & Delaunay** (geometry-based tessellations): Use
        `st_intersection()` to *cut* these geometries at the CONUS
        boundary. These tessellations create boundaries that extend well
        beyond CONUS; intersection actually trims them.
    -   **Hexagon & Square grids** (regular grids): Use `st_filter()` to
        keep only tiles that touch CONUS. These grids are already
        regularly aligned; filtering just discards edge tiles without
        modifying them.

```{r}
voroni_crop<-voronoi_county %>% st_intersection(conus_simp)
tri_crop<-tri_county %>% st_intersection(conus_simp)


grid_crop<-county_grid %>% st_filter(conus_simp)

hex_crop<-county_hex %>% st_filter(conus_simp)
```

::: callout-tip
## Why Different Cropping Methods?

**The distinction matters for your spatial analysis:**

-   **`st_intersection()` on Voronoi/Delaunay**: Produces
    edges/boundaries that align precisely with the CONUS border. Tiles
    at the edge are cut, not discarded. This is geometrically accurate
    but creates irregular edge shapes.

-   **`st_filter()` on hexagon/square grids**: Keeps only complete
    regular tiles that touch CONUS, discarding edge tiles. The boundary
    is a stair-step pattern following the regular grid.

**In Q3**, when you compare these tessellations, you'll notice that
Voronoi and Delaunay have denser tile coverage (e.g., \~2800–6200 tiles)
because they extend to the exact boundary. Hexagons and squares have
fewer tiles (e.g., \~1900–2000) because they're bound to a regular grid.
**This difference affects spatial statistics**: denser tessellations
reveal finer-grained patterns; coarser tessellations show broader
regional structure. This is part of the Modifiable Areal Unit Problem
(MAUP).
:::

### Step 1.6

The last step is to plot your tessellations. We don't want to write out
5 ggplots

Instead, lets make a function that takes an `sf` object as *arg1* and a
character string as *arg2* and returns a ggplot object showing *arg1*
titled with *arg2*.

------------------------------------------------------------------------

The form of a function is:

```{r}
#| echo: true
#| eval: false

tes_func = function(arg1, arg2) {
  
  ggplot(arg1)+
    geom_sf(fill="white",color="navy",size=0.2)+
    labs(title=arg2, caption = paste("Number Features:",nrow(arg1)))+
    theme_void()
  
}



```

------------------------------------------------------------------------

For this function:

-   The name can be anything you chose, *arg1* should take an `sf`
    object, and *arg2* should take a character string that will title
    the plot

-   In your function, the code should follow our standard `ggplot`
    practice where your data is *arg1*, and your title is *arg2*

-   The function should also enforce the following:

    -   a `white` fill

    -   a `navy` border

    -   a `size` of 0.2

    -   \`theme_void\`\`

    -   a caption that reports the number of features in *arg1*

        -   You will need to paste character stings and variables
            together.

```{r}
tes_func = function(arg1, arg2) {
  
  ggplot(arg1)+
    geom_sf(fill="white",color="navy",size=0.2)+
    labs(title=arg2, caption = paste("Number Features:",nrow(arg1)))+
    theme_void()
  
}

```

### Step 1.7

Use your new function to plot each of your tessellated surfaces and the
original county data (5 plots in total):

```{r}
tes_func(arg1 = county_conus,arg2 = "CONUS County Boundaries")
tes_func(arg1 = voroni_crop,arg2 = "Voronoi Tessellation")
tes_func(arg1 = tri_crop, arg2 = "Triangulated Tessellation")
tes_func(arg1 = grid_crop,"Gridded Coverage")
tes_func(arg1 = hex_crop,"Hexagonal Coverage")
```

# Question 2:

In this question, we will write out a function to summarize our
tessellated surfaces.

### Step 2.1

First, we need a function that takes a `sf` object and a `character`
string and returns a `data.frame`.

For this function:

-   The function name can be anything you chose, *arg1* should take an
    `sf` object, and *arg2* should take a character string describing
    the object

-   In your function, calculate the area of `arg1`; convert the units to
    km^2^; and then drop the units

-   Next, create a `data.frame` containing the following:

    1.  text from *arg2*

    2.  the number of features in *arg1*

    3.  the mean area of the features in *arg1* (km^2^)

    4.  the standard deviation of the features in *arg1*

    5.  the total area (km^2^) of *arg1*

-   Return this `data.frame`

```{r}
#Function
func_sum<-function(arg1,arg2){
  areakm2<-drop_units(set_units(st_area(arg1),"km2"))
  
  data.frame(
    type= arg2,
    count=length(areakm2),
    mean=round(mean(areakm2),2),
    sd=round(sd(areakm2),2),
    total=sum(areakm2)
  )
}

```

### Step 2.2

Use your new function to summarize each of your tessellations and the
original counties.

```{r, echo=FALSE}
func_sum(county_id,"Original County")
func_sum(voroni_crop, "Voronoi")
func_sum(tri_crop, "Triangulation")
func_sum(grid_crop, "Grid")
func_sum(hex_crop, "Hexigonal")

```

### Step 2.3

Multiple `data.frame` objects can bound row-wise with `bind_rows` into a
single `data.frame`

For example, if your function is called `sum_tess`, the following would
bind your summaries of the triangulation and voroni object.

```{r, eval = T}
summary_table_Q2<-bind_rows(
  func_sum(county_id,"Original County"),
  func_sum(tri_crop, "Triangulation"),
  func_sum(voroni_crop, "Voronoi"),
  func_sum(grid_crop, "Grid"),
  func_sum(hex_crop, "Hexigonal")
)
```

### Step 2.4

Once your 5 summaries are bound (2 tessellations, 2 coverage's, and the
raw counties) print the `data.frame` as a nice table using
`knitr`/`kableExtra`/`flextable` or your preferred table formatting
package.

```{r, echo = FALSE}

library(flextable)

flextable(summary_table_Q2) %>% 
  set_header_labels(
    type="Type",
    count="Count",
    mean="Mean Area (km²)",
    sd= "Standard Deviation",
    total="Total Area (km²)"
  ) %>% autofit()

```

### Step 2.5

Comment on the traits of each tessellation. Be specific about how these
traits might impact the results of a point-in-polygon analysis in the
contexts of the modifiable areal unit problem and with respect
computational requirements.

> Grid and Hexagon tessellations have a standard deviation of 0. This
> makes sense since since all these cells are uniform compared to the
> other methods which have variation. This uniformity could help with
> computation. Triangulation has a much higher count than the other
> methods, but a smaller mean and smaller total area. This method could
> be advantageous here because this method creates more triangles where
> points are dense which could provide a better local representation.
> Voronoi also adapts to where points are, but it has a higer standard
> deviation than triangulation.

------------------------------------------------------------------------

# Question 3: Dam Distribution Analysis & Tessellation Selection

The data we will analyze in this lab are from the U.S. Army Corps of
Engineers National Dam Inventory (NID). This dataset documents \~91,000
dams in the United States and includes attributes such as design
specifications, risk level, age, and purpose.

Dam management agencies face a critical question: **Where are dams
actually clustered, and how do I visualize that clustering in a way that
supports infrastructure planning decisions?** The answer depends on your
choice of geographic boundaries—that is, your tessellation. Different
tessellations will reveal different patterns of dam concentration. Your
job in this question is to:

1.  Quantify dam distributions across multiple tessellation methods
2.  Understand how the **modifiable areal unit problem (MAUP)** creates
    different narratives from the same underlying data
3.  **Select and justify the tessellation that best supports management
    decisions** about dam clustering

For the remainder of this lab we will analyze the distributions of these
dams using point-in-polygon analysis.

### Step 3.1

In the tradition of this class - and true to data science/GIS work - you
need to find, download, and manage raw data. THe NID dataset is
available as a CSV download from the USACE website
[here](https://nid.sec.usace.army.mil/nid/#/downloads). Grab the CSV.

-   Return to your RStudio Project and read the data in using the
    `readr::read_csv`
    -   Use `janitor::clean_names()` to standardize column names to
        snake_case (tidy data convention)
    -   After reading the data in, be sure to remove rows that don't
        have location values (`!is.na()`)
    -   Convert the `data.frame` to a `sf` object by defining the
        coordinates and CRS
    -   Transform the data to a CONUS AEA (EPSG:5070) projection -
        matching your tessellation
    -   Filter to include only those within your CONUS boundary

::: callout-note
## Data Cleaning with janitor

The NID CSV has messy column names (all caps with spaces). The
`janitor::clean_names()` function converts them to clean snake_case
(`LONGITUDE` → `longitude`, `DAM NAME` → `dam_name`). This is a
lightweight utility that saves time and prevents typos in downstream
code. It's part of data wrangling best practice.
:::

```{r}
#| message: false
# Prefer a local NID CSV named `nation.csv`; otherwise download to that filename
nid_url <- "https://nid.sec.usace.army.mil/api/nation/csv"
dir.create("data", showWarnings = FALSE)
nid_dest <- "data/nation.csv"
if (file.exists(nid_dest)) {
  message("Using local NID file: ", nid_dest)
} else {
  tryCatch({
    message("Downloading NID CSV to: ", nid_dest)
    download.file(nid_url, nid_dest, quiet = FALSE, mode = "wb")
  }, error = function(e) {
    message("NID download failed: ", e$message)
  })
}

# read the CSV (will error if download failed and file missing)
if (!file.exists(nid_dest)) stop("NID CSV not found at ", nid_dest)
dams <- readr::read_csv(nid_dest, skip =1, show_col_types = FALSE) |> 
  janitor::clean_names()

usa <- AOI::aoi_get(state = "conus") %>% 
  st_union() %>% 
  st_transform(5070)

dams2 <- dams %>% 
  filter(!is.na(latitude)) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>% 
  st_transform(5070) %>% 
  st_filter(usa)
```

### Step 3.2

Following the in-class examples develop an efficient point-in-polygon
function that takes:

-   points as `arg1`,
-   polygons as `arg2`,
-   The name of the id column as `arg3`

The function should make use of spatial and non-spatial joins, sf
coercion and `dplyr::count`. The returned object should be input `sf`
object with a column - `n` - counting the number of points in each tile.

::: callout-important
## Point-in-Polygon Function Design Notes

1.  **CRS Matching**: Ensure `points` and `polygons` are in the same CRS
    before calling `st_join()`. If they differ, use `st_transform()` to
    align them first.

2.  **Empty Polygons**: `st_join(...) %>% count(id)` creates one row per
    intersection found. Polygons with *zero* intersections will appear
    in the result with `n = 0` **only if** you use a *left join*
    (default for `st_join()`). If a polygon has no points, make sure
    your result still includes it; if not, use
    `st_join(...) %>% right_join(...)` or initialize counts explicitly.

3.  **Sparse vs. Dense Matrices**: In this step you're not creating
    matrices. However, in Q6.1 you'll use
    `st_touches(..., sparse = FALSE)` to build neighbor matrices
    explicitly. Here in Q3, `st_join()` handles the spatial predicate
    internally (`st_intersects` by default), so you won't set
    sparsity—just know that `st_join()` scales well to large datasets.
    For troubleshooting performance: if your function runs slowly with
    91,000 dams, check whether your tessellation polygons are
    topologically simple (use `st_is_valid()`).
:::

```{r}
st_crs(dams2)==st_crs(usa)


pinp_func <- function(points, polygon, name) {
  st_join(polygon, points) %>%
    st_drop_geometry() %>%
    count(.data[[name]]) %>%
    setNames(c(name, "n")) %>%
    left_join(polygon, by = name) %>%
    st_as_sf()
}


```

### Step 3.3

Apply your point-in-polygon function to each of your five tessellated
surfaces where:

-   Your points are the dams
-   Your polygons are the respective tessellation
-   The id column is the name of the id columns you defined.

```{r}

county_pinp<-pinp_func(points = dams2,polygon = county_id,name = "id")
voronoi_pinp<-pinp_func(points = dams2,polygon = voroni_crop,"id")
tri_pinp<-pinp_func(dams2,tri_crop,"id")
hex_pinp<-pinp_func(dams2,hex_crop,"id")
grid_pinp<-pinp_func(dams2,grid_crop,"id")



```

### Step 3.4

Lets continue the trend of automating our repetitive tasks through
function creation. This time make a new function that **extends your
previous plotting function**.

For this function:

-   The name can be anything you chose, *arg1* should take an `sf`
    object, and *arg2* should take a character string that will title
    the plot

-   The function should also enforce the following:

    -   the fill aesthetic is driven by the count column `n`

    -   the col is `NA`

    -   the fill is scaled to a continuous `viridis` color ramp

    -   `theme_void`

    -   a caption that reports the number of dams in *arg1* (e.g.
        `sum(n)`)

        -   You will need to paste character stings and variables
            together.

```{r}

pinp_plot<-function(data,title){
  ggplot(data)+
    geom_sf(aes(fill=n),color=NA)+
    scale_fill_viridis_c()+
    labs(title = title,caption = paste0("Total Number of Dams:",sum(data$n)))+theme_void()
}
```

### Step 3.5

Apply your plotting function to each of the 5 tessellated surfaces with
Point-in-Polygon counts:

```{r}
pinp_plot(county_pinp,"County Boundary")
pinp_plot(voronoi_pinp,"Voronoi Tessellation")
pinp_plot(tri_pinp,"Triangulated Tessellation")
pinp_plot(hex_pinp,"Hexigonal Tessellation")
pinp_plot(grid_pinp,"Gridded Coverage Tessellation")

countyplot3<-pinp_plot(county_pinp,"County Boundary")
vorplot3<-pinp_plot(voronoi_pinp,"Voronoi Tessellation")
triplot3<-pinp_plot(tri_pinp,"Triangulated Tessellation")
hexplot3<-pinp_plot(hex_pinp,"Hexigonal Tessellation")
gridplot3<-pinp_plot(grid_pinp,"Gridded Coverage Tessellation")
```

### Step 3.6

**Policy Analysis: Tessellation and Decision-Making**

Now that you have dam counts across all five tessellation methods,
compare the results. Specifically:

**a.** Create a side-by-side visualization (faceted or small multiples)
showing dam clustering under each of the five tessellations. Use
`patchwork` or `cowplot` to arrange the 5 plots you created in Step 3.5
for direct visual comparison.

**b.** Identify **2–3 geographic regions** where tessellation choice
produces substantially *different* patterns (e.g., one method shows a
clear hotspot while another shows diffuse distribution). For each
region, describe the pattern difference and explain *why* the
tessellation geometry creates divergent results. Reference the
characteristics from your Q2 summary (number of tiles, mean area,
standard deviation).

**c.** **Which single tessellation would you recommend to a dam
operations manager seeking to understand geographic clustering?**
Justify your choice by addressing: - Does the tessellation align with
natural or management decision-making units (e.g., river basins, states,
hydrologic regions)? - Does it reveal meaningful patterns without
over-fragmenting the landscape or creating artificial boundaries? - How
does MAUP influence your recommendation—are other tessellations equally
valid, just for different questions? - How did simplification affect
your choice? Would a coarser or finer simplification (different `keep`
percentage) change your recommendation? - What does "best supports
management decisions" mean concretely in your context—does it matter
more that hot spots are obvious, or that boundaries align with
jurisdictions?

Your answer should be **3–4 paragraphs** and grounded in the actual
visualizations and patterns you observe from your tessellation
comparisons.

```{r}
library(patchwork)
library(cowplot)

plot_grid(countyplot3,vorplot3,triplot3,hexplot3,gridplot3,ncol = 2,align = "hv")



```

> There is a spot near the Canadian border, near where Montanta is, that
> produces different patterns of brightness. For Gridded and Hexigonal
> coverage this spot is almost unnoticeable because of the uniform tiles
> for this method (sd=0 from table). So this hotspot is averaged over
> the grid/hexagon making it not look obvious. This compares to Voronoi
> and the original county boundary that is bright because this
> tessellation adapts to the points. Triangulated is in the middle of
> these two because it produces a high tile count, but it is still
> changing based on the points. Another area of interest is a spot
> in/near North Caroilina where the Voronoi and county have highlighted
> brightly, but it is not noticable on the triangulated tessellation,
> and this spot is more muted on hexigonal and gridded. This is a
> similar effect to before because equal area grids average out over a
> larger area while the voronoi captures a smaller area making the
> effect look more intense.
>
> The single tessellation I would recommend to a dam operations manager
> seeking to understand geographic clustering would be hexagonal. This
> type of tessellation is unbiased from man made boundaries and evenly
> distributed over the geographic region so the overly break up the
> land' like triangulation and Voronoi can. This will still show
> geographic hotspots, but will not over inflate a pattern in a smaller
> area. This type of tessellation is also easier for computational
> purposes, which may be advantageous if looking at the wide extent of
> the United States. Simplification can affect the choice of
> tessellation because a coarser simplification can limit the quality of
> details, while a finer simplification can preserve more detail, but
> each has a tradeoff with computational performance.
>
> \
> Another tessellations that could be just as important would be using
> the original county polygon. Not every state or even county operate
> the same, so it may be important to understand dam distribution in the
> context of human made boudaries. This would still show some of the
> same hotspots as Voronoi, but Voronoi does not consider political
> boundaries. Counties are spatially different, especially when
> comparing the size of them in the East to the West. The counties in
> the East are smaller than in the West so this may affect spatial
> analysis.

------------------------------------------------------------------------

# Question 4

The NID provides a comprehensive data dictionary
[here](https://files.hawaii.gov/dbedt/op/gis/data/nid_dams_data_dictionary.htm#Purposes).
In it we find that dam purposes are designated by a character
[code](https://files.hawaii.gov/dbedt/op/gis/data/nid_dams_data_dictionary.htm#Purposes).

These are shown below for convenience (built using knitr on a data.frame
called `nid_classifier`):

```{r}
#| echo: false
#| eval: true

nid_classifier = data.frame(matrix(c('I' , 'Irrigation',
'H' , 'Hydroelectric',
'C' , 'Flood Control',
'N' , 'Navigation',
'S' , 'Water Supply',
'R' , 'Recreation',
'P' , 'Fire Protection',
'F' , 'Fish and Wildlife',
'D' , 'Debris Control',
'T' , 'Tailings',
'G' , 'Grade Stabilization',
'O' , 'Other'), ncol = 2, byrow = T)) %>% 
  setNames(c("abbr", "purpose"))
```

```{r}
#| echo: false
#| eval: true

knitr::kable(nid_classifier,              
        caption = "NID 2019: Dam Purposes",
        format.args = list(big.mark = ",")) %>% 
    kableExtra::kable_styling("striped", full_width = TRUE, font_size = 14)
```

-   In the data dictionary, we see a dam can have *multiple* purposes.

-   In these cases, the purpose codes are concatenated in order of
    decreasing importance. For example, `SCR` would indicate the primary
    purposes are *Water Supply*, then *Flood Control*, then
    *Recreation.*

-   A standard summary indicates there are over 400 unique combinations
    of dam purposes:

```{r, echo = TRUE, eval = TRUE}
unique(dams2$purposes) %>% length()
```

-   By storing dam codes as a concatenated string, there is no easy way
    to identify how many dams serve any one purpose... for example where
    are the hydro electric dams?

------------------------------------------------------------------------

To overcome this data structure limitation, we can identify how many
dams serve each purpose by splitting the PURPOSES values (`strsplit`)
and tabulating the unlisted results as a data.frame. Effectively this is
double/triple/quadruple counting dams bases on how many purposes they
serve:

```{r}
#| echo: false
#| eval: false
# create a vector of all characters in your purpose and unlist 
dam_freq <- strsplit(dams2$purposes, split = "") %>%
  unlist() %>% 
  table() %>% 
  as.data.frame() %>% 
  setNames(c("abbr", "count")) %>% 
  left_join(nid_classifier) %>% 
  mutate(lab = paste0(purpose, "\n(", abbr, ")"))
```

The result of this would indicate:

```{r}
#| echo: false
#| eval: false

ggplot() + 
  geom_col(data = dam_freq, aes(x = reorder(lab, -count), y = count, fill = purpose), alpha = .8) + 
  labs(x = "", y = "# Dams", title = "Number of Dams serving each purpose")  +
  scale_fill_viridis_d()+ 
  theme_linedraw() +
  theme(axis.text.x = element_text(angle = 45, vjust = .9, hjust = 1)) + 
  theme(legend.position = 'none')
```

### Step 4.1

-   Your task is to create point-in-polygon counts for at *least* 4 of
    the above dam purposes:

-   You will use `grepl` to filter the complete dataset to those with
    your chosen purpose

-   Remember that `grepl` returns a boolean if a given pattern is
    matched in a string

-   `grepl` is vectorized so can be used in `dplyr::filter`

------------------------------------------------------------------------

For example:

```{r}
#| echo: true
#| eval: true
# Find flood control dams in the first 5 records:
dams2$purposes[1:5]
grepl("F", dams2$purposes[1:5])
```

------------------------------------------------------------------------

For your analysis, choose *at least* four of the above codes, and
describe why you chose them. Then for each of them, create a subset of
dams that serve that purpose using `dplyr::filter` and `grepl`

Finally, use your `point-in-polygon` function to count each subset
across your **cropped tessellation from Q1.4** (e.g., `hexagon_clipped`
or `square_clipped`). (County)

```{r}
#4 dam purposes with grepl
irr_sub <- dams2 %>%
  filter(grepl("I", purposes))

fire_sub <- dams2 %>%
  filter(grepl("P", purposes))

ws_sub <- dams2 %>%
  filter(grepl("S", purposes))

rec_sub <- dams2 %>%
  filter(grepl("R", purposes))


#pinp counts for conus from 1.4

I_pinp <- pinp_func(irr_sub, county_id, "id")
P_pinp<-pinp_func(fire_sub,county_id, "id")
S_pinp<-pinp_func(ws_sub,county_id, "id")
R_pinp<-pinp_func(rec_sub,county_id,"id")


```

### Step 4.2

-   Now use your plotting function from Q3 to map these counts on the
    **cropped tessellation**.

-   *But!* you will use `gghighlight` to **only** color those tiles
    where the count (*n*) is greater then the
    (`mean + 1 standard deviation`) of the set

-   Since your plotting function returns a `ggplot` object already, the
    `gghighlight` call can be added "`+`" directly to the function.

-   The result of this exploration is to highlight the areas of the
    country with the most

```{r}
I_plot<-pinp_plot(I_pinp, "Irrigation Dams")+gghighlight(n>(mean(n)+sd(n)))
P_plot<-pinp_plot(P_pinp, "Fire Protection Dams")+gghighlight(n>(mean(n)+sd(n)))
S_plot<-pinp_plot(S_pinp, "Water Supply Dams")+gghighlight(n>(mean(n)+sd(n)))
R_plot<-pinp_plot(R_pinp, "Recreation Dams")+gghighlight(n>(mean(n)+sd(n)))


plot_grid(I_plot,P_plot,S_plot,R_plot,ncol = 2,align = "hv")+plot_annotation(title="Hot Spot Dam Purposes in CONUS")
```

### Step 4.3

Comment on the geographic distribution of dams you found. Does it make
sense? How might the tessellation you chose impact your findings? How
does the distribution of dams coincide with other geographic factors
such as river systems, climate, ect?

> Geographic distribution varies from these dams. For example fire
> protection dams are almost absent from the east, but very prevalent in
> the Rocky Mountain area extending from Colorado to Montana. This makes
> sense because this area is more prone to wildfires. Water supply is
> more common in the West of the Mississippi. Irrigation Dams are common
> in the West, this makes sense because they don't have enough water
> flowing and need to store it. I might have expected more in some
> places in the Midwest and California, but the locations make sense.
> Recreation Dams are scattered throughout the US. There is an absence
> in the Montana area, likely because these dams are used for other
> purposes like fire protection and water supply. For this tessillation
> choice, counties are much more spread out in the West than in the
> East. This can impact these findings because Western counties can be
> more readily noticed than the small counties. More data points could
> be going into a larger county area making it seem more extreme.
> Eastern states could have the same number of dams in the same area,
> but it will look different because county area is smaller.

**Optional: Connect to Q6** — For purposes you chose in Q4, hypothesize
in 1–2 sentences: *"Will \[irrigation/hydroelectric\] dams show the same
LISA hotspots as all dams in Q6, or different patterns? Why?"* After
completing Q6, revisit this hypothesis.

# Question 5: Spatial Correlation of Dam Attributes — Introduction to Spatial Regression

Dam characteristics—such as **age, surface area, storage capacity, and
purpose**—vary geographically. Understanding the spatial structure of
these relationships is foundational to spatial regression modeling
(upcoming topics). In this question, you'll examine whether dam
attributes exhibit **spatial dependence** and assess whether neighboring
tiles have correlated characteristics.

::: callout-note
## Spatial Regression Motivation (Week 3-1 Foundation)

Standard regression assumes observations are **independent**. But
spatial data often violates this: dams in the Pacific Northwest likely
share design and management practices, leading to **spatially correlated
residuals**. This spatial dependence can bias confidence intervals and
inflate significance tests in standard OLS regression.

The solution involves **spatial lag models** (Spatial Autoregressive,
SAR) or **spatial error models** (SEM) that explicitly model dependence
via neighbor relationships. This lab asks: *Do your tessellation
neighborhoods reveal spatial structure in dam attributes?*

**For this question, use your cropped tessellation from Q1.4** (your
recommended tessellation from Q3, clipped to the CONUS boundary). You'll
examine whether dam attributes (age, size, count) are spatially
autocorrelated—evidence that neighboring tiles have similar values,
violating independence assumptions.
:::

### Step 5.1

**Aggregate Dam Attributes by Tile**

Using your **cropped tessellation from Q1.4** (your recommended grid
from Q3, clipped to CONUS boundary), compute **mean and variance** of
key dam attributes within each tile. The NID dataset typically includes:

-   **Age**: `year_completed` or similar (compute mean age per tile)
-   **Size**: `surface_area_acres` or `normal_pool_elevation` (compute
    mean per tile)
-   **Purpose codes**: `purposes` field (compute dominant purpose per
    tile; count of dams by purpose)

Aggregate as follows:

```{r}
#| eval: false
# Suggested workflow (write your own implementation):
# 1) spatial join dams to your cropped tessellation
# 2) group by tile id and summarize n_dams, mean_year_completed, mean_surface_area
# 3) join summaries back to geometry and preserve zero-dam tiles
# 4) inspect distributions before mapping
```

**Produce visualizations** showing spatial patterns:

-   Map 1: Mean dam age by tile (use viridis scale)
-   Map 2: Mean surface area by tile
-   Map 3: Dam count by tile (you already have this from Q3)

Do you see geographic clustering of old vs. new dams? Large vs. small
dams?

```{r}
agr_dam<-st_join(hex_crop,dams2) %>% group_by(id) %>% summarise(
  n_dams = n(),
  mean_age = mean(year_completed,na.rm=T),
  mean_SA =mean(surface_area_acres, na.rm=T)
) %>% ungroup() %>% st_as_sf()


agr_plot1<-ggplot(agr_dam)+geom_sf(aes(fill=mean_age),color=NA)+scale_fill_viridis_c()+ labs(title= "Mean Dam Age by Tile",fill="Mean Age")+theme_void()


agr_plot2<-ggplot(agr_dam)+geom_sf(aes(fill=mean_SA),color=NA)+scale_fill_viridis_c()+ labs(title= "Mean Surface Area by Tile",fill="Mean Surface Area")+theme_void()

agr_plot3<-ggplot(agr_dam)+geom_sf(aes(fill=n_dams),color=NA)+scale_fill_viridis_c()+ labs(title= "Dam Count by Tile",fill="Number of Dams")+theme_void()


agr_dam_county<-st_join(county_id,dams2) %>% group_by(id) %>% summarise(
  n_dams = n(),
  mean_age = mean(year_completed,na.rm=T),
  mean_SA =mean(surface_area_acres, na.rm=T)
) %>% ungroup() %>% st_as_sf()

agr_cplot1<-ggplot(agr_dam_county)+geom_sf(aes(fill=mean_age),color=NA)+scale_fill_viridis_c()+ labs(title= "Mean Dam Age by Tile",fill="Mean Age")+theme_void()


agr_cplot2<-ggplot(agr_dam_county)+geom_sf(aes(fill=mean_SA),color=NA)+scale_fill_viridis_c()+ labs(title= "Mean Surface Area by Tile",fill="Mean Surface Area")+theme_void()

agr_cplot3<-ggplot(agr_dam_county)+geom_sf(aes(fill=n_dams),color=NA)+scale_fill_viridis_c()+ labs(title= "Dam Count by Tile",fill="Number of Dams")+theme_void()





```

```{r}

plot_grid(agr_plot1,agr_plot2,agr_plot3,ncol = 1)+plot_annotation(title= "Spatial Dam Patterns Hexagonal Tesselation")

plot_grid(agr_cplot1,agr_cplot2,agr_cplot3,ncol = 1)+plot_annotation(title= "Spatial Dam Patterns by County")


```

> Older dams cluster in the eastern U.S while newer dams are in the
> West. Larger mean surface area is clustered by the great lakes area,
> but not very noticeable elsewhere. The dam count is especially
> noticeable along the Mississippi River Basin.

### Step 5.2

**Neighbor Correlation: Local Spatial Variation**

For your tessellation’s neighbor matrix, compute the spatial lag of dam
attributes: the average value of an attribute in a tile’s neighbors.
This reveals whether spatial neighbors have similar dam attributes—a key
signal that observations are not independent (violating OLS
assumptions).

::: callout-warning
## Critical: NA Handling in Spatial Lag Calculations

Many tiles may have **zero dams** (NAs for attributes like
mean_year_completed, mean_surface_area). When you compute the spatial
lag via matrix multiplication, these NAs propagate and can cause `cor()`
to fail with "no complete element pairs."

**Solution: Impute before spatial lag calculation**

1.  Replace NAs in the attribute with the **global mean** (mean across
    all tiles, ignoring NAs)
2.  Compute the spatial lag on the imputed data
3.  When correlating, use **only tiles that had original data** (not
    imputed)

This preserves statistical integrity: neighbors' contributions are
computed correctly, but you only correlate where real observations
exist.
:::

```{r}
#| eval: false
# Suggested workflow (write your own implementation):
# 1) build vectors for count, age, and size attributes by tile
# 2) handle NAs explicitly before matrix multiplication
# 3) compute spatial lag as W_std %*% attribute
# 4) correlate each attribute with its lag
# 5) report values and classify as weak/moderate/strong with your own thresholds
```

**Summarize in a table**:

```{r}
#| eval: false
# Create a compact summary table with:
# - attribute name
# - correlation with spatial lag
# - qualitative strength label
# Then add a 1-2 sentence interpretation below the table.
```

```{r, warning=F}
#Std weights

standardize_weights <- function(geometry_sf) {

  if (any(!st_is_valid(geometry_sf))) {
    geometry_sf <- st_make_valid(geometry_sf)
  }
  
  W <- st_touches(geometry_sf, sparse = FALSE)
  diag(W) <- 0 
  
  W_std <- W / rowSums(W)
  W_std[rowSums(W) == 0, ] <- 0  
  
  return(W_std)
}


W_std<-standardize_weights(agr_dam)
stopifnot(is.matrix(W_std))
stopifnot(all(rowSums(W_std) >= 0))


##Lags
spatial_lag_func <- function(attribute_vector, weight_matrix, original_na_mask = NULL) {
  
  if (is.null(original_na_mask)) {
    original_na_mask <- is.na(attribute_vector)
  }
  
  global_mean <- mean(attribute_vector, na.rm = TRUE)
  attribute_imputed <- ifelse(is.na(attribute_vector), 
                              global_mean, 
                              attribute_vector)
  
  # Compute spatial lag
  spatial_lag <- weight_matrix %*% attribute_imputed
  spatial_lag <- as.vector(spatial_lag)
  
  return(list(
    lag = spatial_lag,
    imputed = attribute_imputed,
    na_mask = original_na_mask
  ))
}

lag_result_n<-spatial_lag_func(agr_dam$n_dams, W_std)
lag_result_age<-spatial_lag_func(agr_dam$mean_age,W_std)
lag_result_SA<-spatial_lag_func(agr_dam$mean_SA,W_std)



cor_n <- cor(lag_result_n$imputed,    lag_result_n$lag, use = "complete.obs")

cor_age <- cor(lag_result_age$imputed[!lag_result_age$na_mask],
               lag_result_age$lag[!lag_result_age$na_mask], use = "complete.obs")


cor_SA <- cor(lag_result_SA$imputed[!lag_result_SA$na_mask],
                lag_result_SA$lag[!lag_result_SA$na_mask], use = "complete.obs")


#Check rq
sum(is.na(lag_result_n$lag))
sum(is.na(lag_result_age$lag))
sum(is.na(lag_result_SA$lag))

#Correlation
# >6 Strong
# 0.3-0.6 Moderate
# <0.3 Weak
#Negative Anticorrelation

classify_strength <- function(cor_val) {
  if (is.na(cor_val)) return(NA)
  if (cor_val > 0.6) return("Strong")
  if (cor_val > 0.3) return("Moderate")
  if (cor_val > 0) return("Weak")
  if (cor_val < 0) return("Negative")
  return("None")
}

summary_df <- data.frame(
  attribute = c("Dam Count", "Mean Age", "Mean Surface Area"),
  correlation = c(cor_n, cor_age, cor_SA)
) %>%
  mutate(
    strength = sapply(correlation, classify_strength),
    correlation = round(correlation, 3)
  )

ftQ5 <- flextable(summary_df) %>%
  set_header_labels(
    attribute = "Attribute",
    correlation = "Correlation with Spatial Lag",
    strength = "Strength"
  ) %>%
  theme_vanilla() %>%
  autofit() 

ftQ5

```

> This table shows that dam count and average age have a strong
> correlation for the spatial clustering while mean surface area has a
> very weak correlation.

### Step 5.3

**Global Question:** Do neighboring tiles resemble each other overall,
or is the pattern weak?

Create a single **Moran's I scatterplot** for **one** dam attribute from
Step 5.2 (for example, mean year completed or dam count). Use
standardized values on both axes: the attribute on the x-axis and the
spatial lag on the y-axis. The plot should show whether points
concentrate along the diagonal or remain widely dispersed.

Your interpretation should focus on the **overall pattern**, not on
specific regions. In particular, explain whether the correlation from
Step 5.2 suggests strong, moderate, or weak spatial dependence and
whether the scatterplot supports that conclusion.

Do **not** create a geographic map here. Save local hotspot mapping for
Q6.

::: callout-tip
## Standardization Details

**Why standardize?** Centering and scaling puts the attribute and
spatial lag on the same scale (mean=0, SD=1), making it easy to identify
quadrants and interpret outliers.

``` r
scale(x, center=TRUE, scale=TRUE)[, 1]
```

Returns a numeric vector (not a matrix). Use `[, 1]` to extract it, or
use `as.vector()` explicitly.
:::

```{r}
#| eval: false
# Suggested workflow (write your own implementation):
# 1) standardize the selected attribute and its spatial lag
# 2) assign HH / LL / HL / LH quadrants by sign of (z, lag_z)
# 3) produce a Moran scatterplot with dashed axes at 0
# 4) report quadrant counts and interpret what dominates
```

```{r}
#LISA Func
assign_lisa_quadrants <- function(attribute_vector, spatial_lag_vector) {
  # Standardize both
  z_attr <- as.vector(scale(attribute_vector))
  z_lag  <- as.vector(scale(spatial_lag_vector))
  
  # Assign quadrants
  quadrant <- case_when(
    z_attr > 0 & z_lag > 0 ~ "HH",
    z_attr < 0 & z_lag < 0 ~ "LL",
    z_attr > 0 & z_lag < 0 ~ "HL",
    z_attr < 0 & z_lag > 0 ~ "LH",
    TRUE ~ NA_character_
  )
  
  return(list( z_attr = z_attr, z_lag = z_lag, quadrant = quadrant))
}



lag_result  <- lag_result_n  

lisa_result <- assign_lisa_quadrants(lag_result_n$imputed, lag_result_n$lag)

moran_data <- data.frame(
  z     = lisa_result$z_attr,
  lag_z = lisa_result$z_lag,
  quad  = lisa_result$quadrant
)

ggplot(moran_data, aes(x = z, y = lag_z, color = quad)) +
  geom_point(size = 2, alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray") +
  scale_color_manual(values = c("HH" = "#d73027", "LL" = "#4575b4",
                                "HL" = "#fee090", "LH" = "#91bfdb"),
                     na.value = "gray80") +
  labs(title = "Moran Scatterplot Number of Dams",
       x = "Standardized attribute", y = "Standardized spatial lag",
       color = "Quadrant") +
  theme_minimal()

```

### Step 5.4

**Global Interpretation of Spatial Patterns**

::: callout-tip
## Writing Guide: Interpreting Spatial Correlation

Use this structure to organize your response:

1.  **State your findings clearly** — cite specific correlation values
    from your Step 5.2 table. Which attributes show the strongest
    clustering?
2.  **Connect to structure** — what does the scatterplot suggest about
    broad spatial organization, without moving into local mapping yet?
3.  **Explain why** — relate to dam history, geology, hydrology, or
    policy at the regional scale
4.  **Analyze patterns** — describe what the Moran scatterplot and
    correlation values reveal about overall dam distributions
:::

Write 2–3 paragraphs (400–600 words) addressing:

**a. Spatial Clustering Patterns (Quantitative + Geographic)**

Which attributes show the strongest spatial lag correlations (from your
Step 5.2 table)? Are correlations \> 0.5 (strong) or 0.2–0.5 (moderate)?

-   **Dam *count***: Often strongly correlated because high dam density
    naturally creates neighbors with many dams.
-   **Dam *age***: Likely shows clustering if dams were built in
    regionally-coherent eras (e.g., 1930s TVA construction, 1960s–70s
    Western expansion).
-   **Dam *size***: May be clustered if geology/hydrology favors larger
    dams in specific regions.

For each attribute, describe whether the clustering seems broad or weak.
Does the correlation make geographic sense? Why should dam
characteristics be similar in neighboring areas?

> As shown in table from 5.2 dam count shows a strong spatial lag
> correlation (hexagonal tessellation) (0.839) indicating that tiles
> with larger dam counts are located near eachother. This is because
> dams are concentrated around areas with people and a river to fill the
> dam. Average dam age also had a strong correlation (0.637). This
> correlation makes sense because as the population expanded further
> west new dams would have to be built for the settlers. Mean surface
> area did not have a strong or moderate corrlation, rather it was found
> to be a weak strength (0.001) with this relationship. This suggests
> that size does not show any strong spatial autocorrelation.

**b. Moran Scatterplot Interpretation (Visualization)**

From your Moran scatterplot, describe the spatial distribution of
points:

-   Do points cluster near the diagonal, suggesting positive spatial
    autocorrelation?
-   Are there many outliers away from the diagonal, suggesting weaker
    dependence or heterogeneous neighborhoods?
-   Does the pattern appear stronger for some attributes than others?

> Points are generally scattered near the diagonal suggesting positive
> spatial autocorrelation. This means that for mean count of damsthe
> high dam count are by other areas with high dam counts and for low dam
> counts the neighbors are also low. These two categories are where the
> majority of the data in the plot is, but there are some points in LH
> and HL. There are a few outliers away from the diagonal, but not
> enough to take away from the overall pattern.

**c. Spatial Structure Implications**

What do these correlation values and Moran scatterplot patterns tell us
about dam infrastructure spatial organization? Given your findings,
write a brief conclusion synthesizing: - Are neighboring tiles genuinely
similar in dam characteristics, or is the clustering weak? - What
management or hydrologic factors drive these patterns? - How do findings
from your Moran analysis here compare to what you observed in Q4 (dam
purposes geography)?

> Spaitial organization is strong for dam age and number of dams and
> weak for dam surface area. Factors that create the spatial
> autocorrelation are likely due to hydrology factors. For example river
> systems are going to have people living near them and along these
> systems dams are impounded to provide water for uses like water supply
> and irrigation, as observed in question 4 distribution. The weak
> correlation found for mean surface area suggests that the size of dam
> varies due to local needs and dam building capabilities. This supports
> the idea regions have different dam infrastructure and uses dependent
> on their needs, and natural availability.

# Question 6: Local Spatial Autocorrelation (LISA) — Mapping Hotspots and Coldspots

**In this question, you'll analyze LISA using your recommended
tessellation from Q3 AND counties as a required comparison unit.** This
contrast (regular tessellation vs administrative boundaries) reveals
MAUP much more clearly than comparing two regular grids.

While global statistics like Moran's I (Q5) test for clustering "on
average," spatial patterns are **rarely uniform**. Some regions have
dense hotspots (e.g., TVA region with concentrated dam infrastructure),
while others are sparse coldspots (remote areas, deserts). **Local
Indicator of Spatial Autocorrelation (LISA)** decomposes global
clustering into local patterns, revealing **where** hotspots and
coldspots occur—critical for targeted management interventions.

In this question, you'll compute LISA using your **cropped recommended
tessellation** (from Q1.4) and **CONUS counties** as the alternative
support. This gives a direct test of how local clustering changes when
analysis units shift from equal-area tiles to administrative units.

::: callout-note
## LISA Concept (Week 3-1 Lectures)

LISA compares each tile's **standardized dam count** ($z_i$) to the
**standardized spatial lag** ($\text{lag}_z$):

$$\text{lag}_z = \frac{1}{w_i} \sum_{j \in N(i)} w_{ij} z_j$$

where neighbors $j \in N(i)$ are adjacent tiles, and $w_{ij}$ =
row-standardized adjacency weight (1 if touching, 0 otherwise).

The resulting **LISA quadrants** reveal: - **High-High (HH)**: High dam
count, high-count neighbors → hotspot - **Low-Low (LL)**: Low dam count,
low-count neighbors → coldspot\
- **High-Low (HL)**: High count, low neighbors → potentially isolated
cluster edge - **Low-High (LH)**: Low count, high neighbors → potential
anomaly or outlier
:::

### Step 6.1

Crop your tessellations to ensure consistent boundaries:

**a.** Use your recommended tessellation from Q3 (e.g., hexagon grid).
Clip it to the CONUS boundary with `st_filter(tessellation, boundary)`.

**b.** Use **counties** as the required alternative comparison unit (no
extra clipping needed if your counties object is already CONUS-only).

**c.** For each clipped tessellation, recompute dam counts using
point-in-polygon joining to ensure correct aggregation within the
cropped geometry.

::: callout-tip
## Cropping Tessellations Correctly

```         
# Recommended tessellation (e.g., hexagon_grid)
hex_cropped <- st_filter(hexagon_grid, usa_boundary)

# Required comparison unit: counties (already CONUS)
county_units <- counties_sf

# Recompute dam counts on cropped grids
# (Use your point_in_polygon function or st_join + count pattern)
hex_dams <- st_join(hex_cropped, dams_sf) %>%
  group_by(id) %>%
  summarise(n = n(), .groups = "drop")

county_dams <- st_join(county_units, dams_sf) %>%
  group_by(id) %>%
  summarise(n = n(), .groups = "drop")
```

**Important**: Cropping changes tile counts and boundary geometries.
Re-aggregate dam counts *after* cropping to match the new tile geometry.
:::

```{r}

# hex_pinp<-pinp_func(dams2,hex_crop,"id")
# county_pinp<-pinp_func(points = dams2,polygon = county_id,name = "id")

hex_counts<-st_filter(hex_pinp,conus_simp) %>% st_make_valid()

county_counts<-st_filter(county_pinp,conus_simp) %>% st_make_valid()

```

### Step 6.2

Compute neighbors and spatial lag for your **recommended tessellation**:

**a.** Create the binary neighbor adjacency matrix using
`st_touches(..., sparse = FALSE)`. Include only tiles within the cropped
boundary.

**b.** Remove self-neighbors: `diag(W) <- 0`

**c.** Row-standardize weights: divide each row by its sum, handling
isolated tiles.

**d.** Compute spatial lag:
$\text{lag} = W_{\text{std}} \times \text{dam\_counts}$

**e.** Standardize both dam counts and spatial lag (z-score
normalization).

::: callout-important
## Validation Checkpoint (Optional but Recommended)

After computing the neighbor matrix, verify it's valid:

```{r}
#| eval: false
# Check 1: Is the matrix symmetric? (for undirected adjacency)
all(neighbor_matrix == t(neighbor_matrix))

# Check 2: Are all rows row-standardized (divide by row sum)?
rowSums(neighbor_weights)  # Should all be 1 (or 0 for isolated tiles)

# Check 3: Are there isolated tiles (no neighbors)?
sum(rowSums(neighbor_weights) == 0)

# If any isolated tiles, document how you'll handle them in spatial lag calculation
```
:::

```{r}

W_std_tess<-standardize_weights(hex_counts)
W_std_county<-standardize_weights(county_counts)

lag_result_tess  <- spatial_lag_func(hex_counts$n, W_std_tess)
lisa_result_tess <- assign_lisa_quadrants(lag_result_tess$imputed, lag_result_tess$lag)

lag_result_county<-spatial_lag_func(county_counts$n, W_std_county)
lisa_result_county<-assign_lisa_quadrants(lag_result_county$imputed,lag_result_county$lag)


hex_counts$z      <- lisa_result_tess$z_attr
hex_counts$lag_z  <- lisa_result_tess$z_lag
hex_counts$lisa   <- lisa_result_tess$quadrant



county_counts$z<-lisa_result_county$z_attr
county_counts$lag_z<-lisa_result_county$z_lag
county_counts$lisa<-lisa_result_county$quadrant




```

### Step 6.3

**LOCAL Question: WHERE specifically are high-high, low-low, high-low,
and low-high clusters?** (Answer: Geographic map showing which regions
belong to which quadrant)

**Two Visualizations:**

**a) Moran Scatterplot (Reference):** - Same as Q5, showing which tile
belongs to which quadrant - Primarily for reference; the interesting
insight is now *geographic*

**b) LISA Geographic Map (PRIMARY OUTPUT):** - Choropleth map of your
tessellation colored by LISA quadrant - HH (red) = hotspots with hotspot
neighbors → TVA region, Colorado Basin, etc. - LL (blue) = coldspots
with coldspot neighbors → Great Basin, deserts, etc. - HL (yellow) =
anomalies: high counts surrounded by sparse tiles (cluster edges) - LH
(cyan) = anomalies: sparse tiles surrounded by high-count neighbors
(rare)

**Key Interpretation:** The LISA map reveals *which specific regions*
exhibit local clustering. This is actionable: dense regions need
intensive management; sparse regions need different strategies. The
quadrant distribution (% of tiles in HH vs. LL) reveals whether the
landscape is dominated by hotspots, coldspots, or mixed.

```{r}
#Lis map
plot_lisa_map <- function(geometry_sf, title = "LISA Map") {
  lisa_colors <- c("HH" = "red", "LL" = "blue",  "HL" = "yellow", "LH" = "cyan", "NA" = "gray90")
  
  ggplot(geometry_sf) +
    geom_sf(aes(fill = lisa), color = NA) +
    scale_fill_manual(values = lisa_colors, na.value = "gray90") +
    labs(title = title) +
    theme_void() +
    theme(legend.position = "bottom")
}


# (a) Moran scatterplot
p_scatter_hex <- ggplot(hex_counts, aes(x = z, y = lag_z, color = lisa)) +
  geom_point(size = 2, alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray") +
  scale_color_manual(values = c("HH" = "#d73027", "LL" = "#4575b4",
                                "HL" = "#fee090", "LH" = "#91bfdb"),
                     na.value = "gray80") +
  labs(title = "LISA Scatter for Hexagonal Tessellation",
       x = "Standardized count", y = "Standardized lag", color = "Quadrant") +
  theme_minimal()

# (b) LISA geographic map
p_map_hex <- plot_lisa_map(hex_counts, "LISA Map Hexagon")

p_scatter_hex + p_map_hex

```

### Step 6.4

**Comparison with required county baseline**:

**a.** Repeat Steps 6.2–6.3 for counties.

**b. REQUIRED DELIVERABLE:** Create side-by-side LISA maps of **your
chosen tessellation (from Q3)** and **counties** in the same figure. Use
identical quadrant labels and color mapping in both panels so visual
differences reflect unit choice, not styling.

**c.** **Quantitative comparison**: For both units, count the number of
polygons in each quadrant. Example output: - Hexagon: HH=15 tiles, LL=23
tiles, HL=5 tiles, LH=2 tiles - Counties: HH=205 counties, LL=1300
counties, HL=240 counties, LH=360 counties - **Interpretation**: Are
quadrant distributions similar (robust LISA pattern) or different (MAUP
sensitive at local scale)?

**d.** Interpret: Which unit reveals clustering more clearly for your
purpose? Do patterns align between your chosen tessellation and
counties, or does MAUP (modifiable areal unit problem) create divergent
local structure?

```{r}
p_scatter_county <- ggplot(county_counts, aes(x = z, y = lag_z, color = lisa)) +
  geom_point(size = 2, alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray") +
  scale_color_manual(values = c("HH" = "#d73027", "LL" = "#4575b4",
                                "HL" = "#fee090", "LH" = "#91bfdb"),
                     na.value = "gray80") +
  labs(title = "LISA Scatter for County Tessellation",
       x = "Standardized count", y = "Standardized lag", color = "Quadrant") +
  theme_minimal()

# (b) LISA geographic map
p_map_county <- plot_lisa_map(county_counts, "LISA Map Counties")


#b map
p_map_hex+p_map_county

#c count
table(hex_counts$lisa)
table(county_counts$lisa)


lis_sum6<-data.frame(
  Tessellation = c("County", "Hexagon"),
  HH = c(756, 525),
  HL = c(232, 111),
  LH = c(317, 170),
  LL = c(1803, 1503)
)

flextable(lis_sum6)

```

> Quadrant distributions are different for the county and hexagonal
> tessellation. The county tessellation has higher counts for
> HH,HL,LH,HL. This is reflected on the map for counties where there are
> more spots, especially to the West. Thus, the MAUP creates divergence
> based on the type of tessellation. Counties vary in size capturing
> different points than hexagon which have a uniform size across the
> country.

### Step 6.5

**Geographic Interpretation and Management Implications**

Write a 3–4 paragraph synthesis addressing:

**a.** **Geographic hotspots and coldspots**: Identify major clustering
regions: - Which dams are in HH hotspots? (E.g., TVA dams, Colorado
River dams, others?) - Which regions are LL coldspots? (Mountains,
deserts, sparse-development areas?) - What management or hydrologic
factors explain these patterns?

> The hotspots for HH are along the Mississippi River basin, the
> multi-state area near where the Ogallala Aquifer is, and also in
> sports near the Rocky Mountains in Colorado and Montanta. The
> Mississippi basin is a wetter area of the U.S. so dams are used to
> control water levels. In Colorado there is a mismatch between water
> and people so water is used for supply and irrigation. In Montana area
> water is used for agriculture and fire protection. Overall, many of
> these areas are important for purposes including flood protection,
> agriculture and water supply. LL coldspots are in areas with less
> people and also in places with a naturally occurring water supply. In
> areas with small populations there is no need for dams, and possibly
> these areas aren't practical to have dams because of a lack of a water
> source. In other areas there is already a natural reservoir or
> groundwater, so there is no need to build.

**Specific region example:** Choose one HH hotspot (e.g., Tennessee
Valley Authority region, Colorado River Basin, California Central
Valley) and describe three factors that explain why TVA dams or Colorado
River dams cluster. Use geographic knowledge or brief web research if
needed.

**b.** **Outliers and anomalies**: Are there notable HL or LH tiles?
What might explain them?

> Most outlier tiles are in the western U.S, more noticeable on the
> county map than the hexagon. For example there are a few HL spots in
> California. This is likely because a populated area needs water supply
> or a dam to support an agriculture area.

**c.** **Unit comparison**: Does analysis unit choice (hexagon vs.
county) affect LISA conclusions? How does MAUP operate at the local
scale?

> Analysis unit choice affects LISA conclusions. The hexagonal
> tessellation has uniform tiles that captures general patterns. This
> compares to county boundary which is fragmented and not uniform across
> the country. This leads to differnt points appearing on the county map
> than the hexagon. Additionally, the counties in the West are larger
> creating more intense color patterns which can make an effect look
> more intense than it really is.

**d.** **Implications for management**: How does LISA output inform dam
operations, reporting, or policy? - Where should multi-state
coordination efforts focus? (HH hotspots) - Which regions have simple
management structures? (LL coldspots) - How do your LISA findings from
Q6 compare to the dam purpose patterns you identified in Q4? (E.g., do
irrigation dams cluster the same way as all dams?) - Where do policies
need special attention to local anomalies? (HL/LH outliers)

> LISA can help show where dam infrastructure already is concentrated.
> This helps reporting, especially as populations grow in places like
> Colorado. While more houses can be built, there isn't always adequate
> water to support these people. Additionally, these hotspots can impact
> interstate boundaries. As a headwater state with a HH hotspot,
> Colorado has to deliver water under different agreements. These dams
> can be used to fulfill these treaties on dry years. Regions with
> simple management structures are LL without many dams usually due to
> small populations. These regions typically have less conflict so
> policy would not have to be as strict here. Dams for different
> purposes, as shown in question 4, are not distributed the same. This
> makes it hard to make policies that apply to all dams across the U.S,
> especially when they allocate water differently across states. For
> example hydroelectic dams won't be as popular in water scarce areas
> while water supply dams will be in these areas. Policies should
> consider place specific needs and inputs making the development as
> local scales more advantageous.
