---
title: 'Lab 2: Distances and Projections'
subtitle: 'The Geometry of Place: Spatial Analysis Skills for Environmental Science'
format:
  html:
    code-fold: true
    toc: true
    include-before-body: ../slides/header.html
    include-after-body: ../slides/footer-annotations.html
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| include: false
knitr::opts_chunk$set(
  message    = FALSE,
  warning    = FALSE,
  comment    = "",
  cache      = FALSE,
  fig.retina = 3
)
```

The spatial analysis skills in this lab — building `sf` objects from
multiple sources, transforming coordinate reference systems, computing
distances to boundaries, and identifying features within threshold zones
— are the same operations you will apply to watersheds, river networks,
gauge networks, and dam inventories throughout this course.

Before those datasets become available, we need the techniques to be
automatic. This lab builds them on a familiar, nationally relevant
dataset: US cities. Every operation you perform here — `st_union`,
`st_combine`, `st_cast`, `st_distance`, `st_transform`, `gghighlight` —
transfers directly to vector water resources data.

The payoff is Q4: a real-world application of these spatial skills to a
live federal policy question that affects more Americans than most
people realize. When you finish, you will have the tools to apply the
same analysis to any spatial threshold problem.

This lab connects directly to recent policy discussions about border
enforcement and the so-called "100-mile" border zone — for example,
shifts in migration policy after the end of the pandemic-era Title 42
expulsions in 2023 have renewed attention on how proximity to the border
affects enforcement, resource allocation, and civil liberties (see
recent coverage:
https://www.reuters.com/world/us/us-ceases-using-title-42-pandemic-order-control-migration-2023-05-11/).

More recent operational discussions involving ICE and border enforcement
through 2025–2026 have emphasized how interior and border-zone
enforcement strategies depend on distance-to-border metrics; readers can
consult summaries on civil-liberties implications (ACLU) or policy
analyses that discuss the "100-mile" enforcement footprint for
background: https://www.aclu.org/know-your-rights/border-zone,
https://www.americanimmigrationcouncil.org/blog/border-patrol-charlotte-atlanta-100-mile-zone/.

------------------------------------------------------------------------

# Set-up {#setup}

## Repository

1.  Navigate to your `csu-523c` repository
2.  Create a new Quarto file called `lab-02.qmd`
3.  Populate its YAML:

``` yaml
---
title: "Lab 2: Distances and Projections"
subtitle: 'ESS 523c: Environmental Data Science Applications'
author:
  - name: Maisie Moline
    email: maimo4@colostate.edu
format: html
---
```

If you dont have a project `_quarto.yml` file, be sure to set the output
directory to `docs`.

## Libraries

Install any missing packages with `install.packages("packagename")`,
then load:

```{r}
#| code-fold: false
pacman::p_load(
  tidyverse, sf, units, tigris, rnaturalearth, gghighlight, ggrepel, flextable
)

options(tigris_use_cache = TRUE)


library(sf)
library(flipbookr)
```

::: callout-tip
Optional hints and worked examples are available if you need extra help:
[hints/lab-02-hints.qmd](hints/lab-02-hints.qmd)
:::

------------------------------------------------------------------------

# Background

In this lab four main spatial skills are covered:

1.  **Building** `sf` objects from R packages and CSV files — Q1
2.  **Manipulating** geometries and coordinate systems — Q2
3.  **Calculating** distances using geometry type selection — Q2
4.  **Visualizing** spatial patterns with `ggplot2`, `gghighlight`, and
    `ggrepel` — Q3 & Q4

::: callout-tip
## Why these skills matter for water resources

Computing distance from a point to a boundary is one of the most common
operations in spatial hydrology: distance from a gauge to a watershed
outlet, distance from a city to the nearest floodplain, distance from a
well to the nearest stream. The mechanics are identical regardless of
what the features represent — master them here and they transfer
everywhere.
:::

------------------------------------------------------------------------

# **Question 1**: Build Your Spatial Dataset {#q1}

For this lab we need three datasets:

1.  CONUS state boundaries (1.1)
2.  North American country boundaries — US, Mexico, Canada (1.2)
3.  All US cities (1.3)

## 1.1 Define a Projection

Distance calculations require a projected coordinate system that
minimizes distance distortion at the scale of CONUS. We will use the
**North America Equidistant Conic**:

```{r}
#| code-fold: false
eqdc <- '+proj=eqdc +lat_0=40 +lon_0=-96 +lat_1=20 +lat_2=60 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs'
```

This PROJ string defines:

| Parameter                | Value      | Meaning                       |
|--------------------------|------------|-------------------------------|
| `+proj=eqdc`             | —          | Equidistant Conic projection  |
| `+lat_0=40`              | 40°N       | Latitude of projection center |
| `+lon_0=-96`             | 96°W       | Central meridian              |
| `+lat_1=20`, `+lat_2=60` | 20°N, 60°N | Standard parallels            |
| `+datum=NAD83`           | —          | North American Datum 1983     |
| `+units=m`               | —          | Output in meters              |

Distortion is minimized between the two standard parallels — 20°N and
60°N encompasses all of CONUS, making this projection well-suited for
national-scale distance analysis.

## 1.2 CONUS State Boundaries

Use `tigris::states()` to get US state boundaries, filter to CONUS, and
transform to the equidistant projection.

::: callout-tip
`tigris` is the modern, CRAN-stable replacement for `USAboundaries`. It
pulls boundaries directly from the US Census Bureau TIGER/Line files.
Load the data with `states(cb = TRUE, progress_bar = FALSE)`. Filter the
state codes (`STUSPS` column) to exclude Alaska, Hawaii, and US
territories, then apply `st_transform(eqdc)` to reproject.
:::

```{r}
conus <- tigris::states(cb=T,progress_bar = FALSE) %>% filter(!NAME %in% c("Alaska","Hawaii","Puerto Rico", "United States Virgin Islands","District of Columbia","Commonwealth of the Northern Mariana Islands","American Samoa","Guam")) %>% st_transform(eqdc)

#DOC or naw




```

## 1.3 North American Country Boundaries

Use `rnaturalearth::ne_countries()` to get country boundaries for the
United States, Canada, and Mexico.

::: callout-tip
`ne_countries()` returns an `sf` object directly when you specify
`returnclass = "sf"`. Filter on the `admin` column to keep only the
three countries you need, then transform to `eqdc`. See
`?rnaturalearth::ne_countries` for available scales and parameters.
:::

```{r}
country_bound<-ne_countries(returnclass = "sf") %>% filter(admin %in% c("United States of America", "Canada","Mexico")) %>% st_transform(eqdc)

```

## 1.4 US City Locations

Go to
[simplemaps.com/data/us-cities](https://simplemaps.com/data/us-cities)
and download the free basic dataset into your `data/` directory.

Read it in with `readr::read_csv()` and explore it — understand what
columns are available before proceeding.

The raw data is tabular — coordinates are stored as numeric columns, not
geometry. Convert it to a spatial object using `st_as_sf()`, specifying
which columns hold X and Y coordinates and what CRS they use. Raw
lat/lon from a CSV is almost always WGS84 (EPSG:4326).

::: callout-tip
After creating the `sf` object, you'll need to remove rows that fall
outside CONUS. The `st_union(states)` geometry provides a boundary to
filter against. See `?sf::st_filter` for how to select geometries that
intersect or fall within a bounding polygon.
:::

```{r}
city<-read.csv("simplemaps_uscities_basicv1.93/uscities.csv")


city_trans <- st_as_sf(city, coords = c("lng", "lat"), crs = 4326) %>%
  st_transform(eqdc)

#conus filt


city_filt<-st_filter(city_trans,st_union(conus))

```

::: callout-important
## Verify all three datasets before proceeding

Run these checks before moving to Q2 — a CRS mismatch here will silently
corrupt every distance calculation downstream.

``` r
# All three must return the same CRS
st_crs(states) == st_crs(countries)
st_crs(states) == st_crs(cities)

# Row counts should be reasonable
nrow(states)    # should be 48
nrow(cities)    # should be ~28,000
```
:::

```{r}
st_crs(conus) == st_crs(country_bound)
st_crs(conus) == st_crs(city_filt)

# Row counts should be reasonable
nrow(conus)    # should be 48
nrow(city_filt)    # should be ~28,000
```

------------------------------------------------------------------------

# **Question 2**: Distance Calculations {#q2}

Here we calculate the distance from each US city to four boundaries:

1.  The US national border (coastline + land borders, resolved)
2.  The nearest state border (internal boundaries preserved)
3.  The Mexican border
4.  The Canadian border

In all cases, since we want distance to a **border** (a line), we must
cast our polygon geometries to `MULTILINESTRING` before computing
distances. The choice of `st_union` vs. `st_combine` determines whether
interior boundaries are dissolved or preserved — a distinction that
matters for questions 2.1 and 2.2.

::: callout-tip
## The geometry type determines the answer

-   `POLYGON` / `MULTIPOLYGON` → distance to the interior is **0**
    (you're inside the area)
-   `MULTILINESTRING` → distance to the **boundary line** — what we
    actually want

``` r
# Wrong: distance to a polygon
st_distance(city_point, state_polygon)   # returns 0 if city is inside

# Right: distance to the border
st_distance(city_point, state_polygon |> st_cast("MULTILINESTRING"))
```
:::

## 2.1 Distance to US National Border (km)

The national border is the **resolved** outer boundary of all states —
internal state lines dissolved, only the coastline and land borders
remain.

::: callout-tip
Use `st_union()` to merge all state polygons into a single boundary
(dissolving internal borders), then cast to `MULTILINESTRING` so
distances are measured to the **border line** rather than interior.
Compute distance using `st_distance()` and convert the result to
kilometers using `set_units()` from the `units` package — this ensures
clean numeric output and correct unit handling downstream.
:::

Store the result as a new column `dist_to_border` in your cities data
frame. Produce a **flextable** of the 5 cities farthest from the
national border — include city name, state, and distance.

```{r}
comb_state<-st_union(conus) %>% st_cast("MULTILINESTRING")



city_filt<-city_filt %>% 
  mutate(
    dist_to_border = st_distance(city_filt,comb_state) %>% set_units("km") %>% drop_units() %>% as.numeric()
  )



city_filt %>%
  st_drop_geometry() %>%
  arrange(desc(dist_to_border)) %>%
  select(city, state_name, dist_to_border) %>%
  head(5) %>%
  flextable() %>% set_header_labels(
    city = "City",
    state_name = "State Name",
    dist_to_border = "Distance to Border (km)"
  )%>% 
  set_caption("Top 5 Cities Farthest from the CONUS border")
```

## 2.2 Distance to Nearest State Border (km)

The nearest state border uses **preserved** internal boundaries — we
want every state line, not just the outer edge. Use `st_combine()` which
keeps all boundaries, then cast to `MULTILINESTRING`.

Store as `dist_to_state`. Produce a flextable of the 5 cities farthest
from any state border.

::: callout-tip
Think carefully about why the answers for 2.1 and 2.2 differ. The city
farthest from the national border might be close to a state line. The
city farthest from any state line must be far from all internal AND
external borders.
:::

```{r}

state_bord<-st_combine(conus) %>% st_cast("MULTILINESTRING")

city_filt<-city_filt %>% 
  mutate(
    dist_to_state = st_distance(city_filt,state_bord) %>% set_units("km") %>% drop_units() %>% as.numeric()
  )


city_filt %>%
  st_drop_geometry() %>%
  arrange(desc(dist_to_state)) %>%
  select(city, state_name, dist_to_state) %>%
  head(5) %>%
  flextable() %>% set_header_labels(
    city = "City",
    state_name = "State Name",
    dist_to_state = "Distance to State Border (km)"
  )%>% 
  set_caption("Top 5 Cities Farthest from a State border")
```

## 2.3 Distance to Mexico (km)

Isolate Mexico from your countries object. Cast to `MULTILINESTRING`.
Compute distance from each city to the Mexican border. Store as
`dist_to_mexico`. Produce a flextable of the 5 cities farthest from
Mexico.

```{r}
Mexico<-country_bound %>% filter(sovereignt == "Mexico") %>% st_cast("MULTILINESTRING")


city_filt<-city_filt %>% 
  mutate(
    dist_to_mexico = st_distance(city_filt,Mexico) %>% set_units("km") %>% drop_units() %>% as.numeric()
  )


city_filt %>%
  st_drop_geometry() %>%
  arrange(desc(dist_to_mexico)) %>%
  select(city, state_name, dist_to_mexico) %>%
  head(5) %>%
  flextable() %>% set_header_labels(
    city = "City",
    state_name = "State Name",
    dist_to_mexico = "Distance to Border (km)"
  )%>% 
  set_caption("Top 5 Cities Farthest from Mexican border")

```

## 2.4 Distance to Canada (km)

Same approach as 2.3, for Canada. Store as `dist_to_canada`. Produce a
flextable of the 5 cities farthest from Canada.

```{r}
Canada<-country_bound %>% filter(sovereignt == "Canada") %>% st_cast("MULTILINESTRING")


city_filt<-city_filt %>% 
  mutate(
    dist_to_canada = st_distance(city_filt,Canada) %>% set_units("km") %>% drop_units() %>% as.numeric()
  )


city_filt %>%
  st_drop_geometry() %>%
  arrange(desc(dist_to_canada)) %>%
  select(city, state_name,dist_to_canada) %>%
  head(5) %>%
  flextable() %>% set_header_labels(
    city = "City",
    state_name = "State Name",
    dist_to_canada = "Distance to Border (km)"
  )%>% 
  set_caption("Top 5 Cities Farthest from Canadian border")

```

------------------------------------------------------------------------

# **Question 3**: Mapping Distance Patterns {#q3}

Now visualize the distance data you computed. You will use `ggplot2` for
all maps, `ggrepel::geom_label_repel()` for non-overlapping city labels,
and `gghighlight` to emphasize subsets.

## 3.1 Reference Map

Create a base reference map showing:

-   North America (US, Canada, Mexico) as filled grey polygons
-   CONUS state boundaries as dashed lines
-   The 10 largest US cities by population as labeled points

::: callout-tip
Use `geom_sf()` for spatial layers. For labels that don't overlap:

``` r
ggrepel::geom_label_repel(
  data = big_cities,
  aes(geometry = geometry, label = city),
  stat = "sf_coordinates",
  size = 3
)
```

`stat = "sf_coordinates"` extracts point coordinates from the geometry
for label placement — required when using `ggrepel` with `sf` objects.
:::

```{r}
t10_city<-city_filt %>% arrange(desc(population)) %>% slice(1:10)



ggplot()+geom_sf(data = country_bound,fill="grey")+geom_sf(data = conus, fill=NA,color="black",linetype="dashed", size=.2)+geom_sf(data=t10_city,color="red")+
  geom_label_repel(
  data = t10_city,
  aes(geometry = geometry, label = city),
  stat = "sf_coordinates",
  size = 3
)+theme_void()

```

## 3.2 Distance to National Border

Color all US cities by their distance to the national border using a
continuous color scale. Highlight and label the 5 cities farthest from
the border.

Your map should: - Use `scale_color_viridis_c()` for the distance
gradient - Show the CONUS outline - Label the 5 farthest cities with
`ggrepel`

```{r}

city_filt_5<-city_filt %>% arrange(desc(dist_to_border)) %>% slice(1:5)

ggplot()+geom_sf(data = comb_state,fill="grey")+geom_sf(data=city_filt, aes(color=dist_to_border))+scale_color_viridis_c(name = "Distance to Border (km)")+
  geom_sf(data = city_filt_5,color="red")+
  geom_label_repel(
  data = city_filt_5,
  aes(geometry = geometry, label = city),
  stat = "sf_coordinates",
  size = 3
)+theme_void()+labs(title = "CONUS Top 5 Cities Furthest from Border")





```

In 2–3 sentences: (1) where are the cities farthest from the national
border? (2) Does the geographic pattern make sense given the shape of
CONUS? (3) Name one water resources application where distance to a
national border would be a meaningful variable.

> (1)The cities that are furthest from the national border are in the
> center of the continental United States (Kansas). (2) This geographic
> pattern makes sense because it is furthest from the borders and in the
> middle of the country. (3) In water resources, an application using
> distance to national border could be used in the context of
> transnational boundaries. States may have to get a certain amount of
> water to a boundary to meet an agreement.

## 3.3 Distance to Nearest State Border

Color all cities by distance to the nearest state border. Highlight and
label the 5 cities farthest from any state line.

```{r}
city_filt_state5<-city_filt %>% arrange(desc(dist_to_state)) %>% slice(1:5)

ggplot()+geom_sf(data=city_filt, aes(color=dist_to_state),size=1)+scale_color_viridis_c(name = "Distance to State Border (km)")+geom_sf(data = conus,fill=NA)+
  geom_sf(data = city_filt_state5,color="red")+
  geom_label_repel(
  data = city_filt_state5,
  aes(geometry = geometry, label = city),
  stat = "sf_coordinates",
  size = 3
)+theme_void()+labs(title = "Top 5 Cities with Furthest Distance from State Border")



```

In 2–3 sentences: (1) how does the pattern differ from 3.2? (2) Which
states appear to have the largest interior areas — do any water
resources features (large river basins, aquifer extents) correlate with
the states where cities can be far from borders? (3) What would a
hydrologist use this distance for?

> \(1\) This pattern is different because now distance is from state
> borders, so larger values are now for state with larger areas. (2)
> States with the largest area include Texas, Montana, Arizona, New
> Mexico. Notable water features include the Ogallala aquifer which
> spans across multiple states. Montana has rivers including the
> Yellowstone and the Missouri. (3) A hydrologist could use this
> distance for applications relating to interstate boundaries such as
> how far a city. For example, determining whether a project falls
> within a zone affecting water allocations under compacts.

## 3.4 Equidistance Zone: Mexico and Canada

Identify cities that are **approximately equidistant** from the Mexican
and Canadian borders — within ±100 km of equal distance from both.

::: callout-tip
Create a variable for the absolute difference between the two distances,
then use `gghighlight()` to emphasize cities where that difference is
less than 100 km. Label the 5 most populous cities in this zone with
`ggrepel`.
:::

```{r}
city_filt<-city_filt %>% mutate(
  abs_dif = abs(dist_to_canada- dist_to_mexico)
)
city_filt_equ<-city_filt %>%filter(abs_dif<100)

city_filt_equ5<-city_filt_equ %>%  arrange(desc(population)) %>% slice(1:5)


ggplot()+geom_sf(data = city_filt_equ, size=0.8,aes(color=abs_dif))+scale_color_viridis_c()+
  gghighlight(abs_dif<100)+theme_void()+geom_sf(data = conus, fill=NA)+geom_sf(data = city_filt_equ5, color="red")+
  geom_label_repel(
  data = city_filt_equ5,
  aes(geometry = geometry, label = city),
  stat = "sf_coordinates",
  size = 3
)+labs(title = "Cities Equidistant from Mexico and Canada (<100 km", color="Distance (km)")

```

In 2–3 sentences: (1) describe where the equidistant zone runs
geographically, (2) does it surprise you that this band cuts through the
interior of the country where it does? (3) What does this zone imply
about the relative influence of Canadian vs. Mexican climate systems on
cities in this corridor?

> (1)These cities run through the central US in the west where Mexico
> and Canada are touching the border of the US. In the east, the cities
> skew downward because Mexico's border is not touching the US so the
> distance needs to account for this gap. (2) This visually makes sense
> when seen on the map because you can see Mexico's border moving
> further away while Canada stays the same. (3). This implies that this
> area may experience both Canadian and Mexican climate systems.

------------------------------------------------------------------------

# **Question 4**: Real-World Application — The 100-Mile Border Zone {#q4}

## Background

Federal regulations give U.S. Customs and Border Protection (CBP)
authority to operate within **100 miles of any US "external boundary"**
— including both land borders and coastlines. Within this zone, CBP
agents can conduct stops and searches without a warrant or reasonable
suspicion. Federal courts have repeatedly upheld this authority.

The American Civil Liberties Union has documented and challenged this
policy. Their analysis is
[here](https://www.aclu.org/other/constitution-100-mile-border-zone).

100 miles ≈ **160 kilometers**. Your distance calculations are already
in kilometers.

## 4.1 Quantify the Zone

Using your `dist_to_border` column, calculate:

-   How many cities fall within the 100-mile zone?
-   What is the total population living in cities within the zone?
-   What percentage of the total US city population does this represent?
-   Does your result match the ACLU's published estimate?

Present your findings as a formatted **flextable**.

::: callout-tip
Filter to cities within 160 km, drop geometry if needed, then summarize:
count rows (`n()`), sum population, and compute the percentage relative
to total city population. You will need to reference total city
population across the ***entire** dataset*, not just filtered rows —
plan your calculation carefully.
:::

```{r}
city_filt<-city_filt %>% mutate(
  total_pop = sum(population)
)

city_160<-city_filt %>% st_drop_geometry() %>% filter(dist_to_border<= 160)

city_sum<-city_160 %>% 
summarise(
  n=n(),
  sum= sum(population),
  percent_of_pop = round((sum/total_pop),2)
) %>% slice(1:1)


city_sum %>% flextable() %>% set_header_labels(
    n = "Number of Cities",
    sum = "Total Population",
    percent_of_pop = "Percentage of the Total US"
  )%>% 
  set_caption("Cities within 100-mile Zone of US Border")



```

In 2–3 sentences: (1) report the percentage and compare it to the ACLU's
estimate, (2) explain why the result might be higher than many people
expect given the geography of the US, (3) name one scenario where a
water resources professional might need to navigate federal
jurisdictional boundaries in a similar way — where a regulatory zone
defined by distance from a physical boundary changes what analysis or
permitting is required.

> \(1\) The ACLU estimates around 2/3 US pop lives within 100 mile zone.
> This is close to 0.65 which is what was calculated in this question.
> (2)This result may be higher than expected because a lot of cities are
> located around water, in this case the ocean.(3) In water resources
> this idea may be relevant for wetland boundaries as defined by WOTUS,
> determining whether a project falls within the jurisdictional buffer
> around wetlands changes the permitting and analysis required under the
> Clean Water Act.

## 4.2 Map the Zone

Create a map highlighting all cities within the 100-mile zone using
`gghighlight`.

-   Color cities by distance to the border using a gradient from
    `"orange"` to `"darkred"`
-   Label the **10 most populous** cities in the zone

```{r}
city_160km<-city_filt %>% filter(dist_to_border<= 160)

city_160_10<-city_filt %>% filter(dist_to_border<= 160) %>% arrange(desc(population)) %>% slice(1:10) 

ggplot()+geom_sf(data = state_bord)+geom_sf(data = city_160km,aes(color=dist_to_border))+gghighlight(dist_to_border<=160)+scale_color_continuous(palette = c("darkred","orange"))+labs(color= "Distance to Border", title="U.S. Cities within 100 mile Border Zone",subtitle = "Top 10 most populous cities in this zone are labeled")+
  geom_sf(data = city_160_10,color="mediumpurple")+
  geom_label_repel(
  data = city_160_10,
  aes(geometry = geometry, label = city),
  stat = "sf_coordinates",
  size = 3
)+theme_void()


```

## 4.3 Most Populous City per State in the Zone

Repeat the map from 4.2, but instead of the 10 most populous overall,
label the **most populous city in each state** that falls within the
zone.

::: callout-tip
``` r
filter(cities, as.numeric(dist_to_border) < 160) |>
  group_by(state_name) |>
  slice_max(population, n = 1)
```
:::

```{r}

pop_city_state<-filter(city_filt, as.numeric(dist_to_border) < 160) |>
  group_by(state_name) |>
  slice_max(population, n = 1)

most_pop<-city_filt %>% filter(state_name %in% city_160km$state_name) %>% group_by(state_name) %>% slice_max(population, n=1)


ggplot()+geom_sf(data = state_bord)+geom_sf(data = city_160km,aes(color=dist_to_border))+gghighlight(dist_to_border<=160)+scale_color_continuous(palette = c("darkred","orange"))+labs(color= "Distance to Border", title="U.S. Cities within 100 mile Border Zone",subtitle = "Most populous city in each state are labeled")+
  geom_sf(data = pop_city_state,color="mediumpurple")+
  geom_label_repel(max.overlaps = 15,
  data = pop_city_state,
  aes(geometry = geometry, label = city),
  stat = "sf_coordinates",
  size = 3
)+theme_void()

setdiff(most_pop$city,pop_city_state$city)

```

In 2–3 sentences: (1) how many states have at least one city in the
zone? (2) Are there states where the most populous city is NOT in the
zone — what does that tell you about that state's geography? (3) How
might the jurisdictional boundary created by this rule create challenges
for environmental monitoring programs that cross state or national
boundaries?

> (1). 35 states have at least one city in the zone. (2) Yes there are
> 13 states that have their most populous city outside fo the zone, for
> example in AZ Phoenix is the most populous city in the state, but
> Tucson is the most populous city in the zone. This is probably because
> a main water supply is located away from the zone. (3) This
> jurisdictional boundary complicates environmental monitoring because
> regulatory authority may change, even when water features span across
> these boundaries.

------------------------------------------------------------------------

# **Question 5**: Interactive Leaflet Exercise — Build an interactive map

This hands-on exercise ties the distance calculations and mapping skills
from Q1–Q4 into a single interactive web map. The goal is a compact,
well-documented Leaflet map with layer controls, informative popups, and
two toggleable thematic layers: (A) the ±100 km equidistance band
(Mexico vs Canada) and (B) the 100-mile (≈160 km) national-border zone.

Learning objectives:

-   Practice turning `sf` results into interactive HTML widgets with
    `leaflet`.
-   Build informative popups that show computed distances and metadata.
-   Implement layer groups and a legend so users can toggle overlays.
-   Consider performance and projection issues when publishing web maps.

## Deliverables

Create an interactive `leaflet` map that includes:

-   **Base layers**: CONUS state boundaries and at least one provider
    tile basemap (e.g., CartoDB, OpenStreetMap)
-   **Data layers** (toggleable via `addLayersControl`):
    -   All cities colored by distance to the national border (use a
        color palette like `colorNumeric("viridis")`; see
        `?leaflet::colorNumeric`)
    -   Cities in the equidistance zone (±100 km Mexico/Canada
        difference) highlighted separately
    -   Cities in the 100-mile border zone highlighted separately
-   **Popups**: When you click a city marker, display relevant
    information: city name, state, population, and distances to borders
-   **Legend**: Show what the color scale represents

::: callout-tip
Key `leaflet` functions: - `leaflet()` to initialize -
`addProviderTiles()` for basemaps - `addCircleMarkers()` or
`addMarkers()` for point data - `addPolygons()` for state boundaries -
`addLegend()` for a color scale legend - `addLayersControl()` to make
layers toggleable

See `?leaflet` and the [Leaflet for R
documentation](https://rstudio.github.io/leaflet/) for syntax and
options. Remember that `leaflet` works with WGS84 coordinates
(EPSG:4326), so you may need to transform your data before plotting.
:::

```{r}
library(leaflet)

cities_leaf <- st_transform(city_filt, 4326)
states_leaf <- st_transform(conus, 4326)
city_160km_leaf <- st_transform(city_160km, 4326)
city_filt_equ_leaf <- st_transform(city_filt %>% 
                                     filter(abs(dist_to_mexico - dist_to_canada) < 100), 4326)

vir_pal <- colorNumeric("viridis", domain = cities_leaf$dist_to_border)

leaflet() %>%
  addProviderTiles(providers$CartoDB) %>%
  addPolygons(
    data = states_leaf,
    fill = FALSE,
    color = "black",
    weight = 1,
    group = "States"
  ) %>%
  addCircleMarkers(
    data = cities_leaf,
    color = ~vir_pal(dist_to_border),
    radius = 3,
    stroke = FALSE,
    fillOpacity = 0.7,
    clusterOptions = markerClusterOptions(),
    popup = ~paste0(
  "<b>", city, ", ", state_id, "</b><br>",
  "<b>Population:</b> ", scales::comma(population), "<br><br>",
  "<b>Distance to Borders:</b><br>",
  "US Border: ", round(dist_to_border, 1), " km<br>",
  "Mexico: ", round(dist_to_mexico, 1), " km<br>",
  "Canada: ", round(dist_to_canada, 1), " km"),
    group = "All Cities"
  ) %>%
  addCircleMarkers(
    data = city_160km_leaf,
    color = "darkred",
    radius = 3,
    stroke = FALSE,
    fillOpacity = 0.9,
    group = "100-mile Zone"
  ) %>%
  addCircleMarkers(
    data = city_filt_equ_leaf,
    color = "orange",
    radius = 3,
    stroke = FALSE,
    fillOpacity = 0.9,
    group = "Equidistant Zone"
  ) %>%
  addLegend(
    "bottomright",
    pal = vir_pal,
    values = cities_leaf$dist_to_border,
    title = "Distance to Border (km)"
  ) %>%
  addLayersControl(
    overlayGroups = c("All Cities", "100-mile Zone", "Equidistant Zone"),
    options = layersControlOptions(collapsed = FALSE)
  )

#Bold header <b> 
#Line Break <br>

```

**Methods note**: In 2–3 sentences, describe the CRS transformation
needed to convert from `eqdc` to `leaflet`-compatible coordinates, and
any design choices you made to keep the map responsive (e.g., filtering,
clustering, or simplifying geometries).

> ESPG data is transformed to WGS84, required for leaflet mapping. I
> used st_transform() to make this transformation. I also clustered the
> city points so they will only break into individual points when
> zooming in.

------------------------------------------------------------------------

# Summary

In this lab you:

-   **Built** spatial datasets from three different sources: a CRAN
    package (`tigris`), a data package (`rnaturalearth`), and a
    downloaded CSV
-   **Transformed** coordinate reference systems and selected a
    projection appropriate for national-scale distance analysis
-   **Manipulated** geometry types — understanding when to use
    `st_union` vs. `st_combine`, and why casting to `MULTILINESTRING` is
    required before computing distances to borders
-   **Computed** four sets of distances using `st_distance()` and stored
    them as labeled unit vectors
-   **Visualized** spatial patterns with continuous color scales,
    `gghighlight`, and `ggrepel`
-   **Built** an interactive web map using `leaflet` to present distance
    data to a general audience

Every one of these operations recurs in the labs ahead — applied to
watersheds, river networks, elevation grids, and gauge networks. The
mechanics are now yours.

------------------------------------------------------------------------

# Rubric

| Question | Topic | Points |
|----------------------------|---------------------|-----------------------|
| Q1 | Build spatial dataset — 3 sources, correct CRS, verification | 25 |
| Q2 | Distance calculations — 4 distances, correct geometry types, flextables | 35 |
| Q3 | Mapping — 4 maps, gghighlight, ggrepel, written interpretations | 20 |
| Q4 | Border zone analysis — table, 2 maps, written interpretation | 35 |
| Q5 | Interactive Leaflet map with layers, popups, legend, and methods note | 35 |
| **Total** |  | **150** |

------------------------------------------------------------------------

# Submission

1.  Render: `quarto render lab-02.qmd`
2.  Push: `git add -A && git commit -m "lab 2 complete" && git push`
3.  Your lab will be live at:
    `https://USERNAME.github.io/csu-523c/lab-02.html`

Submit this URL to Canvas. Add the project to your personal website with
a brief description of what you built — it counts toward end-of-semester
extra credit.
